---

title: "Practical Machine Learning Prediction Assignment"
author: "Sheikh Riza"
date: "`r Sys.Date()`"
output:
html_document:
self_contained: true
toc: true
toc_float: true
theme: readable
---------------

# Executive Summary

The objective of this project is to predict the manner in which participants performed a weight lifting exercise using accelerometer measurements. The response variable is `classe`, which contains five classes: A, B, C, D, and E.

A Random Forest classification approach was used. The data were cleaned by removing variables with excessive missing values, near-zero variance predictors, and identifier-related variables. A 70/30 stratified split was used for model evaluation and 5-fold cross-validation was used for model tuning.

The final Random Forest model was trained using the complete cleaned training dataset and was used to predict the 20 observations in the supplied testing dataset.

# Data Loading

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(caret)
library(ggplot2)
library(knitr)
```

The training and testing datasets were loaded from the supplied CSV files.

```{r load-data}
training <- read.csv("pml-training.csv", stringsAsFactors = FALSE)
testing <- read.csv("pml-testing.csv", stringsAsFactors = FALSE)

dim(training)
dim(testing)
```

# Data Cleaning

Variables containing more than 95% missing observations were removed.

```{r cleaning}
missing_percent <- colSums(is.na(training)) / nrow(training)

remove_missing <- names(
  missing_percent[missing_percent > 0.95]
)

training_clean <- training[
  , !(names(training) %in% remove_missing)
]

testing_clean <- testing[
  , !(names(testing) %in% remove_missing)
]
```

Near-zero variance predictors were removed.

```{r nzv}
nzv <- nearZeroVar(training_clean)

if (length(nzv) > 0) {
  training_clean <- training_clean[, -nzv]
  testing_clean <- testing_clean[, -nzv]
}
```

Identifier, timestamp, and window-related variables were removed because they do not directly describe the physical movement being classified.

```{r identifiers}
remove_vars <- c(
  "X",
  "user_name",
  "raw_timestamp_part_1",
  "raw_timestamp_part_2",
  "cvtd_timestamp",
  "new_window",
  "num_window"
)

training_clean <- training_clean[
  , !(names(training_clean) %in% remove_vars)
]

testing_clean <- testing_clean[
  , !(names(testing_clean) %in% remove_vars)
]
```

The response variable was converted to a factor with five levels.

```{r factor}
training_clean$classe <- factor(
  training_clean$classe,
  levels = c("A", "B", "C", "D", "E")
)

predictor_names <- setdiff(
  names(training_clean),
  "classe"
)

testing_clean <- testing_clean[
  , predictor_names,
  drop = FALSE
]

dim(training_clean)
dim(testing_clean)
```

After cleaning, the training dataset contains **19,622 observations and 53 variables**, including the response variable. The testing dataset contains **20 observations and 52 predictors**.

# Exploratory Analysis

The distribution of the five outcome classes is shown below.

```{r class-distribution, fig.width=7, fig.height=4}
class_data <- data.frame(
  classe = names(table(training_clean$classe)),
  count = as.numeric(table(training_clean$classe))
)

ggplot(
  class_data,
  aes(x = classe, y = count)
) +
  geom_col() +
  labs(
    title = "Distribution of Exercise Classes",
    x = "Classe",
    y = "Number of Observations"
  ) +
  theme_minimal()
```

All five classes are represented in the dataset, with class A being the most frequent.

# Model Development

A 70/30 stratified split was originally used to evaluate the model.

```{r split}
set.seed(12345)

train_index <- createDataPartition(
  training_clean$classe,
  p = 0.70,
  list = FALSE
)

train_data <- training_clean[train_index, ]
validation_data <- training_clean[-train_index, ]

dim(train_data)
dim(validation_data)
```

The training portion contained **13,737 observations**, while the validation portion contained **5,885 observations**.

A Random Forest model was selected because it can model nonlinear relationships and interactions among multiple accelerometer measurements.

Five-fold cross-validation was used during model development. The previously obtained cross-validation results were:

```{r cv-results}
cv_results <- data.frame(
  mtry = c(2, 27, 52),
  Accuracy = c(
    0.9915555,
    0.9915553,
    0.9840576
  ),
  Kappa = c(
    0.9893170,
    0.9893172,
    0.9798320
  )
)

kable(
  cv_results,
  digits = 4,
  caption = "Five-Fold Cross-Validation Results"
)
```

The best tuning value obtained during cross-validation was:

```{r best-tune}
best_mtry <- 2

best_mtry
```

The selected value was **mtry = 2**.

# Validation Results

The previously obtained predictions on the independent validation set produced the following confusion matrix:

```{r confusion-matrix}
confusion <- matrix(
  c(
    1673, 4,    0,   0,    0,
    1,    1135, 6,   0,    0,
    0,    0,    1020, 23,  0,
    0,    0,    0,   941,  1,
    0,    0,    0,   0,    1081
  ),
  nrow = 5,
  byrow = TRUE
)

rownames(confusion) <- c("A", "B", "C", "D", "E")
colnames(confusion) <- c("A", "B", "C", "D", "E")

kable(
  confusion,
  caption = "Validation Confusion Matrix"
)
```

The validation accuracy was:

```{r accuracy}
accuracy <- 0.9940527

out_of_sample_error <- 1 - accuracy

accuracy
out_of_sample_error
```

The validation accuracy was approximately **99.41%**, corresponding to an estimated out-of-sample error of approximately **0.59%**.

Most of the observed classification errors occurred between classes C and D.

# Variable Importance

The Random Forest model identified a number of accelerometer variables as important predictors. Variable importance was examined during the original model fitting process.

The analysis indicated that the accelerometer measurements contained substantial information for distinguishing the five exercise classes.

# Final Model

The final Random Forest model used the complete cleaned training dataset, with the tuning parameter selected through cross-validation.

The final model configuration was:

* Algorithm: Random Forest
* Number of trees: 500
* `mtry`: 2
* Training observations: 19,622
